Dynamics of Laterally Propagating Flames in X-ray Bursts. I. Burning Front Structure

### 2020-02-04 xrb/submit/paper.tex
suybmitted version
A few basic changes to address some comments by the referee

Small-domain studies of convective burning in XRBs preceding flame
development have been done in two-dimensions \citep{lin:2006,xrb,xrb2}
and three-dimensions \citep{xrb3d}.  These calculations used low Mach
number methods, which approximate the hydrodynamics equations to filter
soundwaves, enabling large timesteps and efficient modeling of
subsonic convection.  While these calculations could not support
the lateral differences needed for flame spreading, they can help
understand the role that convection plays in distributing the initial
burning products vertically throughout the neutron star atmosphere as well
as the nature of any turbulence the burning front might encounter as it propagates
through the atmosphere.

@@

To measure the propagation rate of the burning front, we first collapse our nuclear
energy generation rate ($\dot{e}_\mathrm{nuc}$) data at each time into a 1D radial
profile by averaging over the vertical coordinate. We then take the peak
$\dot{e}_\mathrm{nuc}$ value across all profiles to provide a fixed reference point.
We define the position of the flame front to be the location ahead of hottest part of
the flame where $\dot{e}_\mathrm{nuc}$ first drops to $< 0.1 \%$ of the global maximum. 
This corresponds roughly to the leading edge of the burning region. Averaging
over the vertical coordinate helps to reduce sensitivity to localized fluid motions, as does
tracking the $0.1 \%$ contour rather than a local maximum. As we see in Figure
\ref{fig:flame_speed}, the flame settles into a state of steady propagation after an initial
transient period spanning $\sim 3$ ms. The position data here are well fitted by a linear
function of time, and the resulting slope gives the velocity of the flame front.

@@

\begin{figure}[t]
\centering
\plotone{plots/results/speedplot_all.pdf}
\caption{\label{fig:flame_speed}The position of the burning front for each simulation run as a function of time. The dashed lines show linear least squares fits for $t \gtrsim 6$ ms.}
\end{figure}

@@

We explore the baroclinic instability by looking at the magnitude of the 
baroclinicity, calculated as
\begin{equation}
    \boldsymbol{\psi} = \frac{1}{\rho^2} \nablab p \times \nablab \rho,
\end{equation}
and shows the misalignment of the local density and pressure gradients
(we also explored this in \citep{Malone2014a}). As we are considering a 2D system, the
component of the baroclinicity we consider here (out of the plane)
reflects the misalignment of the fields in the plane of the
simulation. Figure~\ref{fig:baroclinicity} shows that the
baroclinicity peaks along the flame front. This baroclinicity
generates vorticity, which in turn entrains material along the surface
of the flame (see also \citealt{cavecchi:2013}). In 3D, this same vorticity
should perturb the flame front and further increase the flame speed
\citep{Cavecchi2019}.

\begin{figure}[t]
\centering
\plotone{plots/results/baroclinicity}
\caption{\label{fig:baroclinicity} Baroclinicity. This plot shows
$\ln \left(\mathbf{\psi}\right)$ at time $t = 0.02~\mathrm{s}$ for the 10/10 simulation.}
\end{figure}

@@

\begin{figure}[t]
\centering
\plottwo{plots/results/vx_vy_129139}{plots/results/u_rho_129139}
\caption{\label{fig:phase_plots} Phase plots at time $t = 0.02~\mathrm{s}$.
\emph{Left}: Phase plot showing the energy generation rate as a function of
the $x$- and $y$-velocities. The black cross shows the location of $u = v = 0$. \emph{Right}: phase plot showing the energy
generation rate as a function of the $x$-velocity and the density.}
\end{figure}

@@

\begin{figure}[t]
\centering
\plotone{plots/results/abar_temp_129139}
\caption{\label{fig:abar_temp} Phase plot of energy generation rate as a function of $\bar{A}$ and temperature at time $t = 0.02~\mathrm{s}$.}
\end{figure}

>>>>

Small-domain studies of convective burning in XRBs preceding flame
development have been done in two-dimensions \citep{lin:2006,xrb,xrb2}
and three-dimensions \citep{xrb3d}.  These calculations used low Mach
number methods, which approximate the hydrodynamics equations to filter
soundwaves, enabling large timesteps and efficient modeling of
subsonic convection.  While these calculations could not support
the lateral differences needed for flame spreading, they can help
us to understand the role that convection plays in distributing the initial
burning products vertically throughout the neutron star atmosphere as well
as the nature of any turbulence the burning front might encounter as it propagates
through the atmosphere.

@@

To measure the propagation rate of the burning front, we first collapse our nuclear
energy generation rate ($\dot{e}_\mathrm{nuc}$) data at each time into a 1D radial
profile by averaging over the vertical coordinate. We then take the peak
$\dot{e}_\mathrm{nuc}$ value across all profiles to provide a fixed reference point.
We define the position of the flame front to be the location to the right of hottest part of
the flame where $\dot{e}_\mathrm{nuc}$ first drops to $< 0.1 \%$ of the global maximum. 
This corresponds roughly to the leading edge of the burning region. Averaging
over the vertical coordinate helps to reduce sensitivity to localized fluid motions, as does
tracking the $0.1 \%$ contour rather than a local maximum. As we see in Figure
\ref{fig:flame_speed}, the flame settles into a state of steady propagation after an initial
transient period spanning $\sim 3$ ms. The position data here are well fitted by a linear
function of time, and the resulting slope gives the velocity of the flame front.

@@

\begin{figure}[t]
\centering
\plotone{speedplot_all.pdf}
\caption{\label{fig:flame_speed}The position of the burning front for each simulation run as a function of time. The dashed lines show linear least squares fits for $t \gtrsim 6$ ms.}
\end{figure}

@@

We explore the baroclinic instability by looking at the magnitude of the 
baroclinicity, calculated as
\begin{equation}
    \boldsymbol{\psi} = \frac{1}{\rho^2} \nablab p \times \nablab \rho,
\end{equation}
and shows the misalignment of the local density and pressure gradients
(we also explored this in \citealt{Malone2014a}). As we are considering a 2D system, the
component of the baroclinicity we consider here (out of the plane)
reflects the misalignment of the fields in the plane of the
simulation. Figure~\ref{fig:baroclinicity} shows that the
baroclinicity peaks along the flame front. This baroclinicity
generates vorticity, which in turn entrains material along the surface
of the flame (see also \citealt{cavecchi:2013}). In 3D, this same vorticity
should perturb the flame front and further increase the flame speed
\citep{Cavecchi2019}.

\begin{figure}[t]
\centering
\plotone{baroclinicity}
\caption{\label{fig:baroclinicity} Baroclinicity. This plot shows
$\ln \left(\mathbf{\psi}\right)$ at time $t = 0.02~\mathrm{s}$ for the 10/10 simulation.}
\end{figure}

@@

\begin{figure}[t]
\centering
\plottwo{vx_vy_129139}{u_rho_129139}
\caption{\label{fig:phase_plots} Phase plots at time $t = 0.02~\mathrm{s}$.
\emph{Left}: Phase plot showing the energy generation rate as a function of
the $x$- and $y$-velocities. The black cross shows the location of $u = v = 0$. \emph{Right}: phase plot showing the energy
generation rate as a function of the $x$-velocity and the density.}
\end{figure}

@@

\begin{figure}[t]
\centering
\plotone{abar_temp_129139}
\caption{\label{fig:abar_temp} Phase plot of energy generation rate as a function of $\bar{A}$ and temperature at time $t = 0.02~\mathrm{s}$.}
\end{figure}
